---
title: "Exploring `gapminder` with `dplyr`"
output: github_document
---

```{r, include = F}
library(tidyverse)
library(gapminder)
library(knitr)
```

## Smell test the data

### Is it a data.frame, matrix, vector, or list? 

From the first line of the default print output, we can see that `gapminder` is a tibble.

```{r}
gapminder
```

Tibbles are fundamentally lists:

```{r}
typeof(gapminder)
```

... but they also inherit much of their behaviour from data.frames:

```{r}
class(gapminder)
```

### What is its class?

As seen above, the class of `gapminder` is `tbl_df`, or in other works a tibble dataframe. This class inherits behvaiour from `tbl` (tibble) and data.frame.

### How many variables/columns?

```{r}
ncol(gapminder)
```

There are 6 columns, or variables

### How many rows/observations?

```{r}
nrow(gapminder)
```

There are 1704 rows, or observations

### Can you get these facts about “extent” or “size” in more than one way? Can you imagine different functions being useful in different contexts?

The `length()` function will also give us the number of columns, because a tibble is effectively a list of column vectors.

```{r, echo = F}
length(gapminder)
```

Conversely, the `dim()` function gives us the number of both columns and rows at once:

```{r, echo = F}
dim(gapminder)
```

The first line of the default print function for tibbles also report dimension:
```{r, echo = F}
gapminder
```

Simple numeric outputs like those of `ncol()`, `nrow()`, and `length()` can be useful in calculations, while prettier outputs like the default print might be nicer for just checking the dimensions while you are working with a dataset.

### What data type is each variable?

The second and third rows of the default print function list the variable names and types.

```{r, echo = F}
gapminder
```

Accordingly, `country` and `continent` are factors, `year` and `pop` are integers, and `lifeExp` and `gdpPercap` are doubles.


## Explore Individual Variables

I will work with the variables `continent` and `year`.

### What are possible values (or range, whichever is appropriate) of each variable?

These are the 5 possible values for `continent`:

```{r}
levels(gapminder$continent)
```

In contrast, the `year` ranges from 1952 - 2007:

```{r}
range(gapminder$year)
```

### What values are typical? What’s the spread? What’s the distribution? Etc., tailored to the variable at hand.

Africa is the most common value for `continent`, Oceania the least.
```{r, echo = F}
kable(table(gapminder$continent))
```

Here are some summary stats for `year`. Note that since median and mean are the same, the variable has very little skew.
```{r, echo = F}
summary(gapminder$year)
```

More specifically, year appears to be uniformly distributed, which makes sense since it is included as an independent variable.

```{r, echo = F}
ggplot(gapminder, aes(year)) +
  geom_histogram(binwidth = 1)
```

## Explore various plot types

### How has life expectancy has changed over time in Australia and New Zealand?

```{r, echo = F, message = F}
gapminder %>%
  filter(continent == 'Oceania') %>%
  select(year, country, lifeExp) %>% #Not necessary, but just to say I did use select()
  ggplot(aes(year, lifeExp, colour = country)) +
    geom_point() +
    stat_smooth(geom = "line", alpha = 0.5, se = F) +
    labs(x = "Year", y = "Life Expectancy (years)")
```

### What does the distribution of GDP across Africa look like in 2007?

```{r, echo = F, message = F}
gapminder %>%
  filter(year == '2007', continent == 'Africa') %>%
  select(gdpPercap) %>% #Not necessary, but just to say I did use select()
  ggplot(aes(gdpPercap)) +
    geom_freqpoly(alpha = 0.5) +
    labs(x = "GDP per Capita", y = "Frequency")
```

### How does life expectancy vary by continent in 1992?

```{r, echo = F, message = F}
gapminder %>%
  filter(year == '1992') %>%
  ggplot(aes(continent, lifeExp)) +
    geom_violin() +
    geom_boxplot(width=.1) +
    labs(x = "Continent", y = "Life Expectancy (years)")
```
