---
title: "Exploring `gapminder` with `dplyr`"
output: github_document
---

```{r load_packages, include = F}
library(tidyverse)
library(gapminder)
library(knitr)
```

## Smell test the data

### Is it a data.frame, matrix, vector, or list?

From the first line of the default print output, we can see that `gapminder` is a tibble.

```{r view_gapminder}
head(gapminder, n = 2) #Same output as `gapminder` with fewer lines
```

Tibbles are fundamentally lists:

```{r typeof_gapminder}
typeof(gapminder)
```

... but they also inherit much of their behaviour from data.frames:

```{r class_gapminder}
class(gapminder)
```

### What is its class?

As seen above, the class of `gapminder` is `tbl_df`, or in other works a tibble dataframe. This class inherits behvaiour from `tbl` (tibble) and data.frame.

### How many variables/columns?

```{r ncol_gapminder}
ncol(gapminder)
```

There are 6 columns, or variables

### How many rows/observations?

```{r nrow_gapminder}
nrow(gapminder)
```

There are 1704 rows, or observations

### Can you get these facts about “extent” or “size” in more than one way? Can you imagine different functions being useful in different contexts?

The `length()` function will also give us the number of columns, because a tibble is effectively a list of column vectors.

```{r length_gapminder}
length(gapminder)
```

Conversely, the `dim()` function gives us the number of both columns and rows at once:

```{r dim_gapminder}
dim(gapminder)
```

The first line of the default print function for tibbles also report dimension:
```{r view2_gapminder}
gapminder
```

Simple numeric outputs like those of `ncol()`, `nrow()`, and `length()` can be useful in calculations, while prettier outputs like the default print might be nicer for just checking the dimensions while you are working with a dataset.

### What data type is each variable?

The second and third rows of the default print function list the variable names and types.

```{r view3_gapminder}
head(gapminder, n = 2) #Same output as `gapminder` with fewer lines
```

Accordingly, `country` and `continent` are factors, `year` and `pop` are integers, and `lifeExp` and `gdpPercap` are doubles.


## Explore Individual Variables

I will work with the variables `continent` and `year`.

### What are possible values (or range, whichever is appropriate) of each variable?

These are the 5 possible values for `continent`:

```{r levels_gapminder}
levels(gapminder$continent)
```

In contrast, the `year` ranges from 1952 - 2007:

```{r range_gapminder}
range(gapminder$year)
```

### What values are typical? What’s the spread? What’s the distribution? Etc., tailored to the variable at hand.

Africa is the most common value for `continent`, Oceania the least.
```{r kable_gapminder}
kable(table(gapminder$continent)) #Pass the output of table() into kable() for prettier output
```

Here are some summary stats for `year` generated using the `summary()` function. Note that since median and mean are the same, the variable has very little skew.
```{r summary_gapminder}
summary(gapminder$year)
```

More specifically, year appears to be uniformly distributed, which makes sense since it is included as an independent variable.

```{r histogram_year}
ggplot(gapminder, aes(year)) +
  geom_histogram(binwidth = 1)
```

## Explore various plot types

### How has life expectancy has changed over time in Australia and New Zealand?

```{r lifeExp_over_time_Oceania, message = F}
gapminder %>%
  filter(continent == 'Oceania') %>% #Only look at Australia and New Zealand
  select(year, country, lifeExp) %>% #Drop the unused variables. Not necessary, but just to say I did use select()
  ggplot(aes(year, lifeExp, colour = country)) + #Note that the colour argument in aes will group the data for us
    geom_point() +
    stat_smooth(geom = "line", alpha = 0.5, se = F) +
    labs(x = "Year", y = "Life Expectancy (years)")
```

Since the mid-70s, Australia appears to have held a slight lead in life expectancy over New Zealand.

### What does the distribution of GDP per capita across Africa look like in 2007?

```{r gdpPercap_Africa, message = F}
gapminder %>%
  filter(year == '2007', continent == 'Africa') %>% #Only look at African countries in 2007
  select(gdpPercap) %>% #Once again not necessary
  ggplot(aes(gdpPercap)) + #These graphs only need an x variable
    geom_freqpoly(alpha = 0.5) +
    labs(x = "GDP per Capita", y = "Frequency")
```

Many African countries have GDP per capita below 2500USD in 2007.

### How does life expectancy vary by continent in 1992?

```{r lifeExp_by_continent, message = F}
gapminder %>%
  filter(year == '1992') %>% #We only care about the year 1992
  ggplot(aes(continent, lifeExp)) +
    geom_violin() +
    geom_boxplot(width=.1) + #I like narrow boxplots inside my violin plots to see quartiles better
    labs(x = "Continent", y = "Life Expectancy (years)")
```

Africa appears to have the greatest variance in life expectancy in 1992, and Europe appears to be binodal.
